Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-08-03 threshold/draft.Rnw
Removed absolute value in squared terms where appropriate

The following lemma expresses the Euclidean distance between any two columns of
the matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 + 
\sigma_2^2 |e_i^Tv_{2} - e_j^Tv_{2}|^2 + \cdots + 
\sigma_k^2 |e_i^Tv_{ki} - e_j^Tv_{k}|^2,
\end{equation}
\end{lemma}
where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
vector consisting of $1$ in position $j$ and zeros otherwise.
(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$; that is, the
$[j,k]$ entry in the matrix $V$.)

@@

Note that each term in the sum is nonnegative and the weights $\sigma_j^2$,
$j=1,2,\ldots,k$ are nonincreasing and defined by the covariance structure of
the data.  Given a correlation threshold $t$, $0<t<1$, Equations \ref{cordist}
and \ref{vdist} together suggest that pairs of vectors can be excluded from
consideration whenever their projected distance is too large.  For example if
$\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 > 2(1-t)$, then we can conclude from
just two scalar values that the column vectors $a_i$ and $a_j$ do not meet the
correlation threshold $t$.
In practice, many pairs of vectors may be pruned in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.

@@

Let $t$ be a given correlation threshold $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders then entries
of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |e_i^Tv_1 -
e_j^Tv_1|^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
longest run of successive points in $P v_1$ within the interval
$\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1} displays the entries
of $v_1$ for the small ``BicatYeast'' example (where $A\in\mathcal{R}^{70\times
419}$) from Section \ref{examples} in increasing order.  The lines in the plot
illustrate the interval associated with the correlation threshold $t=0.95$. The
quantity $\ell$ is obtained by rolling the interval across all the points in
the plot and counting the maximum number of points that lie within the
interval.  With respect to the ordering $P$, $\ell$ represents the biggest
index difference that pairs of indices corresponding to correlated vectors
above the threshold can exhibit.

>>>>

The following lemma expresses the Euclidean distance between any two columns of
the matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} - e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{ki} - e_j^Tv_{k})^2,
\end{equation}
\end{lemma}
where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
vector consisting of $1$ in position $j$ and zeros otherwise.
(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$; that is, the
$[j,k]$ entry in the matrix $V$.)

@@

Note that each term in the sum is nonnegative and the weights $\sigma_j^2$,
$j=1,2,\ldots,k$ are nonincreasing and defined by the covariance structure of
the data.  Given a correlation threshold $t$, $0<t<1$, Equations \ref{cordist}
and \ref{vdist} together suggest that pairs of vectors can be excluded from
consideration whenever their projected distance is too large.  For example if
$\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 > 2(1-t)$, then we can conclude from
just two scalar values that the column vectors $a_i$ and $a_j$ do not meet the
correlation threshold $t$.
In practice, many pairs of vectors may be pruned in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.

@@

Let $t$ be a given correlation threshold $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders then entries
of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 (e_i^Tv_1 -
e_j^Tv_1)^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
longest run of successive points in $P v_1$ within the interval
$\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1} displays the entries
of $v_1$ for the small ``BicatYeast'' example (where $A\in\mathcal{R}^{70\times
419}$) from Section \ref{examples} in increasing order.  The lines in the plot
illustrate the interval associated with the correlation threshold $t=0.95$. The
quantity $\ell$ is obtained by rolling the interval across all the points in
the plot and counting the maximum number of points that lie within the
interval.  With respect to the ordering $P$, $\ell$ represents the biggest
index difference that pairs of indices corresponding to correlated vectors
above the threshold can exhibit.
